---
title: "Colexification data and phylogenetic trees"
subtitle: "First look at D, s2, and alpha for selected pairs"
date: today
author: "Thomas Brochhagen"
format: html
code-fold: true
bibliography: ../bibliography.bib
---

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(readr)
library(ggplot2)
library(tidyr)
```



```{r}
summarize_tbl_physig <- function(psig_colex){
        tmp_no_mcmc <- psig_colex[ !(psig_colex$tree %in% c("MCMC (summary)", "MCMC (posterior)")), ];
        tmp_no_mcmc$tree <- ifelse(tmp_no_mcmc$tree == "Glottolog (original)", "Original", ifelse(tmp_no_mcmc$tree == "Glottolog (exponential)", "Exponential", ifelse(tmp_no_mcmc$tree == "Glottolog (ultrametric)", "Ultrametric", tmp_no_mcmc$tree)));
        tmp_no_mcmc$error[ is.na(tmp_no_mcmc$error) ] <- "-";
        tmp_no_mcmc$D <- sprintf("%s%.2f%s", ifelse(tmp_no_mcmc$D.p0 < 0.05,"**",""), tmp_no_mcmc$D, ifelse(tmp_no_mcmc$D.p0 < 0.05,"**",""));
        tmp_no_mcmc$D.p0 <- sprintf("%s%.2g%s%s", ifelse(tmp_no_mcmc$D.p0 < 0.05,"**",""), tmp_no_mcmc$D.p0, ifelse(tmp_no_mcmc$D.p0 < 0.01, "^⭑⭑^", ifelse(tmp_no_mcmc$D.p0 < 0.05, "^⭑^", "")), ifelse(tmp_no_mcmc$D.p0 < 0.05,"**",""));
        tmp_no_mcmc$D.p1 <- sprintf("%.2g%s", tmp_no_mcmc$D.p1, ifelse(tmp_no_mcmc$D.p1 < 0.01, "^⭑⭑^", ifelse(tmp_no_mcmc$D.p1 < 0.05, "^⭑^","")));
        tmp_no_mcmc$PGLMM.s2 <- sprintf("%s%.2f%s", ifelse(tmp_no_mcmc$PGLMM.s2.p0 < 0.05, "**", ""), tmp_no_mcmc$PGLMM.s2, ifelse(tmp_no_mcmc$PGLMM.s2.p0 < 0.05, "**", ""));
        tmp_no_mcmc$PGLMM.s2.p0 <- sprintf("%s%.2g%s%s", ifelse(tmp_no_mcmc$PGLMM.s2.p0 < 0.05, "**", ""), tmp_no_mcmc$PGLMM.s2.p0, ifelse(tmp_no_mcmc$PGLMM.s2.p0 < 0.01, "^⭑⭑^", ifelse(tmp_no_mcmc$PGLMM.s2.p0 < 0.05, "^⭑^", "")), ifelse(tmp_no_mcmc$PGLMM.s2.p0 < 0.05, "**", ""));
        tmp_no_mcmc$IG10.alpha <- sprintf("%.2f", tmp_no_mcmc$IG10.alpha);
        tmp_no_mcmc$IG10.alphaWarn <- ifelse( tmp_no_mcmc$IG10.alphaWarn == 0, "-", "at limit");
        tmp_no_mcmc$D[ tmp_no_mcmc$D == "NANANA" ] <- "-";
        tmp_no_mcmc$D.p0[ tmp_no_mcmc$D.p0 == "NANANANA" ] <- "-";
        tmp_no_mcmc$D.p1[ tmp_no_mcmc$D.p1 == "NANA" ] <- "-";
        tmp_no_mcmc$PGLMM.s2[ tmp_no_mcmc$PGLMM.s2 == "NANANA" ] <- "-";
        tmp_no_mcmc$PGLMM.s2.p0[ tmp_no_mcmc$PGLMM.s2.p0 == "NANANANA" ] <- "-";
        tmp_no_mcmc$IG10.alpha[ tmp_no_mcmc$IG10.alpha == "NA" ] <- "-";
        tmp_no_mcmc$IG10.alphaWarn[ is.na(tmp_no_mcmc$IG10.alphaWarn) ] <- "";
        tmp_no_mcmc$family[ tmp_no_mcmc$family == "World" ] <- "***World***";
        return(knitr::kable(tmp_no_mcmc, 
                row.names=FALSE, 
                col.names=c("Family", "Tree", "Errors?", "*D*", "*p*(*D*=0)", "*p*(*D*=1)", "*s2*", "*p*(*s2*=0)", "*α*", "*α* warning?"),
                align="r",
                caption='Various estimates of the phylogenetic signal when using the **Glottolog trees** with original ("Original") branch length (i.e., all branches have length 1.0), using an exponential ("Exponential") branch length and a ultrametric ("Ultrametric") one, as well as the **@jaeger:2018 trees** (where available), and the **@zhang+etal:2019 for Sino-Tibetan**.'))
}
```

```{r}
draw_coverage_map <- function(df, c1, c2){
        df_sub_colex <- df %>% 
                        filter(Concepticon_ID.xo == c1,
                        Concepticon_ID.yo == c2)


        mp <- map_data('world')
        ggplot() +
                geom_polygon(data = mp, aes(x = long, y = lat, group=group)) +
                geom_point(data = drop_na(df_sub_colex), aes(x = longitude, y = latitude, col = colex)) +
                theme_minimal(base_size = 13) +
                theme(axis.title.x=element_blank(),
                axis.text.x=element_blank(),
                axis.ticks.x=element_blank(),
                axis.title.y=element_blank(),
                axis.text.y=element_blank(),
                axis.ticks.y=element_blank(),
                legend.position = c(0.9,0.1),
                legend.background = element_rect(fill='grey',
                                                size=0.25)
                )
}
```

```{r, warning=FALSE, message=FALSE}
df_geo  <- read_csv('../data/glottolog/languages_and_dialects_geo.csv') %>%
        select(glottocode, macroarea, latitude, longitude) %>%
        rename(Glottocode = glottocode)

df_colex <- read_csv('../data/lexibank/lexibank-colex-unif.csv') %>%
                left_join(df_geo) %>%
                mutate(colex = ifelse(colex == 0, 'Absent', 'Present'))
```

### Preamble

We have 19168 unique meaning pairs that, in principle, one could estimate the phylogenetic signal for. We probably will want to set a threshold for the minimal number of languages that a meaning pair has to appear in s.t. it is included in the analysis^[For a rough impression: for such a threshold set at at least 100, we have 13075 meaning pairs; at least 500 yields 3794; at least 1000 yields 1528; 1500 yields 369; and so on.], and possibly another for the minimal number of times it is colexified because otherwise there's nothing to estimate (see "group (iii)" below). 

For now --to design the entire analysis pipeline and for first impressions-- I selected six meaning pairs from three "groups":

(i) Two of the meaning pairs that colexify most often across languages:
*  (i.i) *upper arm* and *lower arm*: $300$ languages/observations with $179$ cases of colexification
* (i.ii) *leg* and *foot*: $298$ observations with $161$ cases of colexification
(ii) Two pairs that are often discussed in the literature in terms of areal colexification effects:
* (ii.i) *fire* and *tree*: $1417$ languages/observations with only $18$ positive cases. As noted by @gast+koptam:2018 this is "is a well-known colexification pattern [...] well-attested in the Sahul area comprising the languages of Australia, New Guinea and surrounding islands.
* (ii.ii) *ear* and *leaf*: $1660$ languages/observations with $8$ positive cases. 
(iii) Two pairs for which we have the most lexical coverage, irrespective of how often they colexify:
* (iii.i) *nose* and *eye*: $1850$ languages/observations with $1$ positive cases.
* (iii.ii) *ear* and *eye*: $1848$ languages/observations with $2$ positive cases.

For the cases in (iii) I obviously didn't have any expectations since it is impossible to say anything sensible about them in terms of phylogenetic signals. I included them for completeness' sake but  only to highlight that going by frequency of lexicalization information alone means that we will end up with many cases like these.

For the cases in (ii) I was uncertain. If this is a bona-fide areal effect then we would expect there to not be much of a phylogenetic signal. However, it can obviously happen that areal effects coincide with phylogenetic effects, with more related languages being spoken in more proximity to one another.

The cases in (i) are probably the most interesting candidates for phylogenetic scrutiny since we have more positive cases of colexification than in groups (ii) and (iii).  

$D$, $s^2$ and $\alpha$ were calculated for all six pairs. The regressions are all intercept-only.


### (i.i) upper and lower arms

The high frequency of colexification of upper and lower arm is the kind of the case for which people criticise Lexibank/CLICS -- should this even be counted as "two distinct meanings"?-- but it's an good testbed for these methods because, a priori, one may not except a (strong, if any) phylogenetic signal. 

#### Map

It makes sense to first look at the distribution of the data on a map. To get a feeling for the coverage we have on whether these two meanings colexify (absent = no colexification; present = colexified).

```{r}
draw_coverage_map(df_colex, 431, 446)
```

This map already hints at the fact that we will not get much family-level diversity from this colexification pair. 


#### Phylogenetic signal

```{r}
load('../cached_results/signals/physig-431-446.RData') #upper - lower arm
summarize_tbl_physig(physig_colex)
```

As suggested by the map above, this pair showcases a challenge that we will face: data on a lexical contrast is either not collected (no one appears to have bothered to ask Austroasiatic or Indo-Euopean speakers if they have different words for upper- and lower-arms); or it is constant, so we get an estimate only for a (likely typologically biased) subsample.

There's not all that much that can be said about this pair based on  $D$, $s^2$ and $\alpha$. 

### (i.ii) legs and feet


#### Map
```{r}
draw_coverage_map(df_colex, 1297, 1301)
```

A priori, this meaning pair looks a little more informative than the previous one. There is more data across more families.

#### Phylogenetic signal
```{r}
load('../cached_results/signals/physig-1297-1301.RData') #fire-tree
summarize_tbl_physig(physig_colex)
```

And, indeed, there is more information to go by for this pair. It's a mixed picture but there's some evidence pointing toward clumpiness. 

### (ii.i) Fire and wood

#### Map
```{r}
draw_coverage_map(df_colex, 221, 906)
```

The main to note is that the pattern is very localized (as expected from the literature), so we will get a constant colexification status for most families.


```{r}
load('../cached_results/signals/physig-221-906.RData') #fire-tree
summarize_tbl_physig(physig_colex)
```

And, indeed, there is little to be said about this pattern either. The only family that gives us some variation in colexification is Pama-Nyungan

### (ii.ii) Ears and leaves

#### Map
```{r}
draw_coverage_map(df_colex, 628, 1247)
```

This case looks very similar to (ii.i)

#### Phylogenetic signal
```{r}
load('../cached_results/signals/physig-628-1247.RData') #fire-tree
summarize_tbl_physig(physig_colex)
```

## Group (iii) remark

These should not be taken seriously because there's barely any positive case of colexification. I ran them because it's cheap; and to get a feeling for the estimates one would get when analyzing this kind of extremely skewed data

### (iii.i) Noses and eyes

#### Map
```{r}
draw_coverage_map(df_colex, 1221, 1248)
```

#### Phylogenetic signal
```{r}
load('../cached_results/signals/physig-1221-1248.RData') #fire-tree
summarize_tbl_physig(physig_colex)
```

### (iii.ii) Ears and eyes

#### Map
```{r}
draw_coverage_map(df_colex, 1247, 1248)
```

#### Phylogenetic signal
```{r}
load('../cached_results/signals/physig-1247-1248.RData') #fire-tree
summarize_tbl_physig(physig_colex)
```



# Some concluding thoughts

* For cases of comparably rare colexifications (fire and wood; ear and leaf) the problem of by-family measures is that they are blind to the uniqueness of a (lack of) pattern in the global context. Then again, maybe it is also telling that --within the families that host them-- they do not appear to be all that clumpy. This may be taken to suggest that this is, indeed, not so much a phylogenetic but an areal effect, as suggested by the literature; 
* We need to discuss how to deal with "group (iii)"-kind of cases where colexification is rare;
* I am not really trained in frequentist methods but the issue of multiple tests does rear its head. 

